---
title: "Texas Power Crisis (February 2021)"
author: "Sofia Sark"
format: html
editor: visual
---

## Initial Set Up

```{r}
#| output: false
# Loading necessary libraries
library(tidyverse)
library(here)
library(terra)
library(sf)
library(tmap)
library(stars)
```

## Create Blackout Mask (night lights data)

```{r}
#| output: false
# Loading in raster (night light) data
feb7_tile5 <- read_stars(here("data/VNP46A1/VNP46A1.A2021038.h08v05.001.2021039064328.tif"))
feb7_tile6 <- read_stars(here("data/VNP46A1/VNP46A1.A2021038.h08v06.001.2021039064329.tif"))
feb16_tile5 <- read_stars(here("data/VNP46A1/VNP46A1.A2021047.h08v05.001.2021048091106.tif"))
feb16_tile6 <- read_stars(here("data/VNP46A1/VNP46A1.A2021047.h08v06.001.2021048091105.tif"))
```

Our data comes in two files (tiles) for each day. First, we must merge them by day.
```{r}
# Merging rasters together, by day
feb7_lights <- st_mosaic(feb7_tile5, feb7_tile6)
feb16_lights <- st_mosaic(feb16_tile5, feb16_tile6)
```

We can find the difference in light intensity between the two days by subtracting them.
```{r}
# Finding difference in light instensity
diff_light <-  feb7_lights - feb16_lights
```

We are only interested in areas where the drop in light intensity is greater than 200. 
```{r}
# Assigning difference in light intensity to new object (to preserve original)
diff_mask <- diff_light

# All differences values less than 200 set as NA
diff_mask[diff_mask < 200] <- NA
```

```{r}
# Vectorizing mask
vector_diff_mask <- diff_mask %>% 
  st_as_sf() %>% 
  st_make_valid()
```

We are only interested in the Houston area, so we will crop our geometries to its area.
```{r}
# Creating a polygon that defines the Houston area based on its coordinates
crop_box <- st_bbox(c(xmin = -96.5, xmax = -94.5, ymin = 29, ymax = 30.5), crs = st_crs(diff_mask)) %>% 
  st_as_sfc()
```

```{r}
# Cropping our light intensity difference mask using our predefined crop polygon
diff_cropped <- st_crop(vector_diff_mask, crop_box)

# Ensuring that our cropping worked by checking for difference in cell number
if(ncell(vector_diff_mask) == nrow(diff_cropped)){
  warning("clipping did not remove cells")
} else {
  print("clipping removed cells")
}
```

```{r}
# Transforming our final light intensity difference to have the NAD83 / Texas Centric Albers Equal Area CRS
diff_final <- st_transform(diff_cropped, crs = 3083)
```

## Exclude highways from cropped blackout mask

Highways could have experienced changes to light intensity that are unrelated to the storm, so we will exclude them from our mask.

Importing roads data
```{r}
#| output: false
# Loading in roads vector data (line object)
# The SQl query allows us to only load highway data (instead of all roads)
highways <- st_read("data/gis_osm_roads_free_1.gpkg",
                 query = "SELECT * FROM gis_osm_roads_free_1 WHERE fclass='motorway'"
) %>% 
  st_transform(3083)  # preemptively changing CRS to match our other data
```
Creating 200 meter buffer around Houston roads.
```{r}
# st_buffer with specified units finds all areas within that distance for us
highways_200m <-  st_buffer(highways, dist = units::set_units(200, "m"))
```

Combining highway data with its buffer
```{r}
highways <- st_simplify(highways, dTolerance = 10)
highways_200m <- st_simplify(highways_200m, dTolerance = 10)

highways_union <- st_union(highways, highways_200m)
# Now we have a single object that includes the highways and 200m around them
```

Now, we find areas that experiences a blackout, but 200m away from highways.
```{r}
blackout_no_highway <- st_difference(diff_final, highways_200m)
```

## Identify the number of homes likely impacted by blackouts

Importing houses data
```{r}
#| output: false
# Loading in houses vector data (multipolygon object)
houses <- st_read(here("data/gis_osm_buildings_a_free_1.gpkg"),
                 query = "SELECT *
FROM gis_osm_buildings_a_free_1
WHERE (type IS NULL AND name IS NULL)
OR type in ('residential', 'apartments', 'house', 'static_caravan', 'detached')") %>% 
  st_transform(3083)  # preemptively changing CRS to match our other data
```

First, we have to create a layer of homes that are only within our `blackout_no_highway` polygon. This is because we want to see *only* difference in light intensity greater than 200 and unaffected by highway light changes.
```{r}
# We use st_contains to find the houses that are fully contained by our blackout polygon
houses_blackout <- st_contains(blackout_no_highway, houses)
```

