---
title: "Texas Power Crisis (February 2021)"
author: "Sofia Sark"
format: html
editor: visual
---

This quarto document contains analyses based in light intensity differences between February 7 and 16, 2021, in Houston, Texas. It compiles remotely-sensed night lights data, geographic features such as roads and buildings, and census income data. This exploration is meant to explore the power crisis that Texas experienced in February 2021 (due to winter storms), and reveal demographic information regarding which homes experienced a blackout.


## Initial Set Up

```{r}
#| output: false
# Loading necessary libraries
library(tidyverse)
library(here)
library(terra)
library(sf)
library(tmap)
library(stars)
```

## Data Cleaning

### Create Blackout Mask (night lights data)

Our night light intensity data comes from the Visible Infrared Imaging Radiometer Suite (VIIRS) onboard the Suomi satellite, which can be accessed from NASA's Level-1 and Atmospheric Archive & Distribution System Distributed Active Archive Center (LAADS DAAC).

We have four data sets in total: two days, and two tiles from each day. This is because Houston splits two of the tiles.
```{r}
#| output: false
# Loading in raster (night light) data
feb7_tile5 <- read_stars(here("data/VNP46A1/VNP46A1.A2021038.h08v05.001.2021039064328.tif"))
feb7_tile6 <- read_stars(here("data/VNP46A1/VNP46A1.A2021038.h08v06.001.2021039064329.tif"))
feb16_tile5 <- read_stars(here("data/VNP46A1/VNP46A1.A2021047.h08v05.001.2021048091106.tif"))
feb16_tile6 <- read_stars(here("data/VNP46A1/VNP46A1.A2021047.h08v06.001.2021048091105.tif"))
```

First, we must merge them by day.
```{r}
# Merging rasters together, by day
feb7_lights <- st_mosaic(feb7_tile5, feb7_tile6)
feb16_lights <- st_mosaic(feb16_tile5, feb16_tile6)
```

We can find the difference in light intensity between the two days by subtracting them.
```{r}
# Finding difference in light instensity
diff_light <-  feb7_lights - feb16_lights
```

We are only interested in areas where the drop in light intensity is greater than 200 nW cm-2sr-1. For the purposes of our analysis, this wil signify a blackout.
```{r}
# Assigning difference in light intensity to new object (to preserve original)
diff_mask <- diff_light

# All differences values less than 200 set as NA
diff_mask[diff_mask < 200] <- NA
```

```{r}
# Vectorizing mask to be able to use stars functions on it
vector_diff_mask <- diff_mask %>% 
  st_as_sf() %>% 
  st_make_valid()
```

We are only interested in the Houston area, so we will crop our geometries to its area.
```{r}
# Creating a polygon that defines the Houston area based on its coordinates
crop_box <- st_bbox(c(xmin = -96.5, xmax = -94.5, ymin = 29, ymax = 30.5), crs = st_crs(diff_mask)) %>% 
  st_as_sfc()
```

```{r}
# Cropping our light intensity difference mask using our predefined crop polygon
diff_cropped <- st_crop(vector_diff_mask, crop_box)

# Ensuring that our cropping worked by checking for difference in cell number
if(ncell(vector_diff_mask) == ncell(diff_cropped)){
  warning("Cropping did not remove cells")
} else {
  print("Cropping removed cells")
}
```
This is our cleaned light intensity difference object! Our final step is to adjust it's CRS.
```{r}
# Transforming our final light intensity difference to have the NAD83 / Texas Centric Albers Equal Area CRS
diff_final <- st_transform(diff_cropped, crs = 3083)
```

### Exclude highways from cropped blackout mask

Highways could have experienced changes to light intensity that are unrelated to the storm, so we will exclude them from our mask.

Our roads data is sourced from Geofabrik (a third party company redistributing OpenStreetMap's available geographic world data). The shapefile was retrieved from Geofabrik’s download sites and the class's amazing teaching team prepared a Geopackage (.gpkg file) containing just the subset of roads that intersect the Houston metropolitan area. 

```{r}
#| output: false
# Loading in roads vector data (line object)
# The SQL query allows us to only load highway data (instead of all roads)
highways <- st_read("data/gis_osm_roads_free_1.gpkg",
                 query = "SELECT * FROM gis_osm_roads_free_1 WHERE fclass='motorway'") %>% 
  st_transform(crs = st_crs(diff_final)) %>%   # preemptively changing CRS to match our other data
  st_union()                                   

# by running st_union, we ensure that all of our roads are clumped into one geometry;
# this can help avoid long running times in the rest of our analysis
```
Creating 200 meter buffer around Houston roads.
```{r}
# st_buffer with specified units finds all areas within that distance for us
highways_200m <-  st_buffer(highways, dist = units::set_units(200, "m")) %>% 
  st_union()

# Comining roads data with its buffer to create single object with highways and 200m around them
highways_union <- st_union(highways, highways_200m)
```

Now, we find areas that experiences a blackout, but 200m away from highways.
```{r}
# st_difference() will return all non-intersecting areas
blackout_no_highway <- st_difference(diff_final, highways_200m)
```

### Identify the number of homes likely impacted by blackouts

Houses data was sourced in the same way as our roads data: it is OpenStreetMap data retrieved using Geofabrick.

```{r}
#| output: false
# Loading in houses vector data (multipolygon object)
# The SQl query allows us to only load in only homes, rather than all buildings
houses <- st_read(here("data/gis_osm_buildings_a_free_1.gpkg"),
                 query = "SELECT * FROM gis_osm_buildings_a_free_1
                          WHERE (type IS NULL AND name IS NULL)
                          OR type in ('residential', 'apartments', 'house', 'static_caravan', 'detached')") %>% 
  st_transform(crs = st_crs(diff_final))  # preemptively changing CRS to match our other data
```

We have to create a layer of homes that are only within our `blackout_no_highway` polygon. This is because we only want to see differences in light intensity greater than 200 nW cm-2sr-1 *and* unaffected by highway light changes.

```{r}
# We use st_contains to find the houses that are fully contained by our blackout polygon
houses_blackout_sgbp <- st_contains(blackout_no_highway, houses)

# Convert sgbp list to an object of booleans (TRUE --> houses are in the blackout area)
houses_blackout_logical <- lengths(houses_blackout_sgbp) > 0 

# Subset only houses in blackout area from all houses data
houses_blackout <- houses[houses_blackout_logical, ]

# Ensuring that our cropping worked by checking for difference in cell number
if(ncell(blackout_no_highway) == ncell(houses_blackout)){
  warning("Failed to extract house data 200m away from highways")
} else {
  print("Successfully extracted only houses affected by blackouts (200m away from highways)")
}
```


### Blackout by Census Tracts

Census tract geometries as well as income data by census tract was sourced from U.S. Census Bureau’s American Community Survey for census tracts in 2019. The ArcGIS "file geodatabase" contains multiple layers of data, so we first extract this information to explore where our data of interest is stored, and then we can retrieve it.

```{r}
# Checking layers within our socioeconomic data
layers <-  st_layers(here("data/ACS_2019_5YR_TRACT_48_TEXAS.gdb"))

# Importing income and geometry layers
income <- st_read(here("data/ACS_2019_5YR_TRACT_48_TEXAS.gdb"), layer = "X19_INCOME")
census_tracts <- st_read(here("data/ACS_2019_5YR_TRACT_48_TEXAS.gdb"), layer = "ACS_2019_5YR_TRACT_48_TEXAS")
```
For this analysis we will use median income. The income layer contains a lot of other data, so we will clean it up to only contain GEOID (which we will need to combine with our spatial data) and median income. We will also clean census tract data to match.

```{r}
# Select only variables of interest from income data
income <- income %>% 
  select("GEOID", "B19013e1") %>%  
  rename(median_income = B19013e1)   # Rename median income column to be more clear

# census_tracts stores geo_ID under a different column names -- fixing it to match income
census_tracts <- census_tracts %>% 
  select(-GEOID) %>%                # It also contained a different, unecessary 'GEOID' column; remove it
  rename(GEOID = GEOID_Data)
```

Our final goal for initial census data cleaning and merging is a data frame with census tract geometries and median income information. We do this by joining the two data frames.
```{r}
# Combine the two layers
income_tracts <- left_join(census_tracts, income, by = "GEOID")
```

We now want to know which census tracts contain houses that experienced a blackout. To do this, we first combine house data with our `income_tracts` data frame.

By utilizing a left join, we can ensure that we keep *all* income tracts, and insert NA values where a house experiencing a blackout was not found.

```{r}
# Ensuring income_tracts has the same CRS
income_tracts <- st_transform(income_tracts, crs = st_crs(houses_blackout))

# Creating a geometry containing only the tracts that have houses with a blackout
houses_tracts <- st_join(income_tracts, houses_blackout, left = TRUE)
```

We can create a new column called `blackout`, and fill it with TRUE/FALSEs based on whether there was a house that experienced a blackout in that census tract (TRUE = there was). We can base this on the `osm_id` column of our `houses_tracts` data frame, because it contains IDs for the blackout houses. If there is no ID (NA), we assume that there is no blackout house in that census tract.

```{r}
# Create new variable 'blackout' where TRUE if osm_id has data
houses_tracts$blackout <- houses_tracts$osm_id > 0

# Replace the NAs in the blackout column with FALSE, for areas with no houses experiencing a blackout
houses_tracts$blackout <- houses_tracts$blackout %>% 
  replace_na(FALSE)
```

Great! Now we have a dataframe containing census tract geometries, their associated median incomes, and information on whether or not any houses in that tract experienced a blackout.

I think I don't need to do this
```{r}
# transforming CRS' to match
census_tracts <- st_transform(census_tracts, crs = st_crs(houses_blackout))

# filtering census_tracts to only include tracts that contain houses w blackouts
census_blackout <- st_filter(census_tracts, houses_blackout)

# trying to find the opposite of census blackout DELETE
census_noblackout_sgbp <- st_disjoint(census_blackout, census_tracts)
census_noblackout_logical <- lengths(census_noblackout_sgbp) < 0
census_noblackout <- census_tracts[census_noblackout_logical,]

# Joining tract data to match census data that experienced blackouts DELETE?
houses_tracts <- st_join(income_tracts, census_blackout, join = st_within)

if(ncell(census_tracts) == ncell(census_blackout)){
  warning("Filtering did not remove cells")
} else {
  print("Filtering successfully removed cells")
}
```


## Maps, Figures and Reflection

The following section contains the following outputs:
- a set of maps comparing night light intensities before and after the first two storms
- a map of the homes in Houston that lost power
- an estimate of the number of homes in Houston that lost power
- a map of the census tracts in Houston that lost power
- a plot comparing the distributions of median household income for census tracts that did and did not experience blackouts
- a brief reflection (approx. 100 words) summarizing your results and discussing any limitations to this study

### Comparing Light Intensities (Map)

```{r}
# Cropping both days of our light data to just the Houston area
# Using previously made crop box
feb7_cropped <- st_crop(feb7_lights, crop_box)
feb16_cropped <- st_crop(feb16_lights, crop_box)

# Plotting light intensity for both days
feb7 <- plot(feb7_cropped,
     main = "Before Storm (2021-02-07)")

feb16 <- plot(feb16_cropped,
     main = "After Storm (2021-02-07)")

night_light_colors <- c("0" = "#0d1166", "100" = "#47497a","200" = "#e4e5f5", "300" = "#f5f5e4", "400" = "white")

feb7_plot <- tm_shape(feb7_cropped) +
  tm_raster(breaks = c(0, 100, 200, 300, 400, 500, Inf),
            palette = night_light_colors,
            title = "Light Intensity (nW cm-2sr-1)") +
  tm_title(text = "February 7, 2021") +
  tm_graticules(alpha = 0.5)

legend_map <-  tm_shape(feb7_cropped) +
  tm_raster(breaks = c(0, 100, 200, 300, 400, 500, Inf),
            palette = night_light_colors,
            title = "Light Intensity (nW cm-2sr-1)") +
  tm_title(text = "February 7, 2021") +
  tm_graticules(alpha = 0.5) +
  tm_layout(legend.only = TRUE)

feb16_plot <- tm_shape(feb16_cropped) +
  tm_raster(breaks = c(0, 100, 200, 300, 400, 500, Inf),
            palette = night_light_colors,
            title = "Light Intensity (nW cm-2sr-1)") +
  tm_title(text = "February 16, 2021") +
  tm_graticules(alpha = 0.5) +
  tm_layout(legend.show = FALSE)

tmap_arrange(feb7_plot, feb16_plot) 
```

### Homes in Houston that Lost Power (Map & Esimate)

**Map**

```{r}
tm_shape(houses_tracts, bbox = houses_blackout) +
  tm_polygons(fill = "#c6b8cf",
              alpha = 0.25,
              border.col = "gray90",
              lwd = 0.75) +
tm_title(text = "Houston, TX: Homes that experienced blackout in February 2021") +
tm_compass() +
tm_scale_bar() +
tm_shape(houses_blackout) +
  tm_symbols(fill = "darkblue",
             border.col = "darkblue",
             alpha = 0.5,
             size = 0.15) 
```

**Estimate**

Our data frame `houses_blackout` only contains houses in areas that we determined experience a blackout. Therefore, to find the number of homes in Houston that experiences a blackout, we can simply count the number of rows in our data frame.
```{r}
houses_estimate <- nrow(houses_blackout)

print(paste("The number of homes that lost power in Houston between February 7th and 16th, 2021, is", houses_estimate, "."))
```

## Blackout and Census Tract

```{r}
tm_shape(census_tracts, bbox = houses_blackout) +
  tm_polygons(fill = "#c6b8cf",
              border.col = "gray90",
              lwd = 0.75) +
tm_shape(census_blackout) +
  tm_polygons(fill = "#482161",
              border.col = "#c6b8cf",
              lwd = 0.75) +
tm_add_legend(labels = c("Experienced blackout", "Did not experience blackout"),
              fill = c("#482161", "#c6b8cf"),
              title = "Census Blocks",
              is.portait = TRUE) +
tm_title("February 2021 Storms: Houston, TX") +
tm_compass() +
tm_scale_bar()
```

a plot comparing the distributions of median household income for census tracts that did and did not experience blackouts

```{r}
tm_shape(houses_tracts) +
  tm_polygons(fill = "blackout")

  tm_shape(census_blackout) +
  tm_polygons(fill = "#482161",
              border.col = "#c6b8cf",
              lwd = 0.75) +
tm_add_legend(labels = c("Experienced blackout", "Did not experience blackout"),
              fill = c("#482161", "#c6b8cf"),
              title = "Census Blocks",
              is.portait = TRUE) +
tm_title("February 2021 Storms: Houston, TX") +
tm_compass() +
tm_scale_bar()
```


```{r}
# census blocks that experiences blackout
# census_blackout

# all census blocks with income data
# income_tracts

# I don't need geometries anymore

income_tracts_df <- income_tracts %>% 
  st_drop_geometry() %>% 
  mutate(median_income = B19013e1)

census_blackout_df <- census_blackout %>% 
  st_drop_geometry()

# joining onto census blackout df will only leave income data for census blocks with blackouts
blackout_income <- inner_join(income_tracts_df, census_blackout_df, by = "NAMELSAD")

# to keep the opposite observations, we can use anti_join
noblackout_income <- anti_join(income_tracts_df, blackout_income, by = "NAMELSAD")


```

```{r}
# plot of median income distribution of houses experiencing blackout
plot1 <- blackout_median <- ggplot(blackout_income, aes(y = median_income)) +
  geom_boxplot() +
labs(title = "Census Tracts Experiencing Blackout")

plot2 <- noblackout_median <- ggplot(noblackout_income, aes(y = median_income)) +
  geom_boxplot() +
labs(title = "Census Tracts No Blackout")

library(patchwork)

plot1 + plot2
```

Questions for Office Hours:
- how do I change color bar/legend as well as color in general for plots showing differences in light intensity
- ^ also, adding long/lat lines possible?
- is estimate just number of rows? and is plot of just dots ok? 
- are my distributions of median income right???