---
title: "Texas Power Crisis (February 2021)"
author: "Sofia Sark"
format: html
editor: visual
---

## Initial Set Up

```{r}
#| output: false
# Loading necessary libraries
library(tidyverse)
library(here)
library(terra)
library(sf)
library(tmap)
library(stars)
```

## Data Cleaning

### Create Blackout Mask (night lights data)

```{r}
#| output: false
# Loading in raster (night light) data
feb7_tile5 <- read_stars(here("data/VNP46A1/VNP46A1.A2021038.h08v05.001.2021039064328.tif"))
feb7_tile6 <- read_stars(here("data/VNP46A1/VNP46A1.A2021038.h08v06.001.2021039064329.tif"))
feb16_tile5 <- read_stars(here("data/VNP46A1/VNP46A1.A2021047.h08v05.001.2021048091106.tif"))
feb16_tile6 <- read_stars(here("data/VNP46A1/VNP46A1.A2021047.h08v06.001.2021048091105.tif"))
```

Our data comes in two files (tiles) for each day. First, we must merge them by day.
```{r}
# Merging rasters together, by day
feb7_lights <- st_mosaic(feb7_tile5, feb7_tile6)
feb16_lights <- st_mosaic(feb16_tile5, feb16_tile6)
```

We can find the difference in light intensity between the two days by subtracting them.
```{r}
# Finding difference in light instensity
diff_light <-  feb7_lights - feb16_lights
```

We are only interested in areas where the drop in light intensity is greater than 200. 
```{r}
# Assigning difference in light intensity to new object (to preserve original)
diff_mask <- diff_light

# All differences values less than 200 set as NA
diff_mask[diff_mask < 200] <- NA
```

```{r}
# Vectorizing mask
vector_diff_mask <- diff_mask %>% 
  st_as_sf() %>% 
  st_make_valid()
```

We are only interested in the Houston area, so we will crop our geometries to its area.
```{r}
# Creating a polygon that defines the Houston area based on its coordinates
crop_box <- st_bbox(c(xmin = -96.5, xmax = -94.5, ymin = 29, ymax = 30.5), crs = st_crs(diff_mask)) %>% 
  st_as_sfc()
```

```{r}
# Cropping our light intensity difference mask using our predefined crop polygon
diff_cropped <- st_crop(vector_diff_mask, crop_box)

# Ensuring that our cropping worked by checking for difference in cell number
if(ncell(vector_diff_mask) == ncell(diff_cropped)){
  warning("clipping did not remove cells")
} else {
  print("clipping removed cells")
}
```

```{r}
# Transforming our final light intensity difference to have the NAD83 / Texas Centric Albers Equal Area CRS
diff_final <- st_transform(diff_cropped, crs = 3083)
```

### Exclude highways from cropped blackout mask

Highways could have experienced changes to light intensity that are unrelated to the storm, so we will exclude them from our mask.

Importing roads data
```{r}
#| output: false
# Loading in roads vector data (line object)
# The SQl query allows us to only load highway data (instead of all roads)
highways <- st_read("data/gis_osm_roads_free_1.gpkg",
                 query = "SELECT * FROM gis_osm_roads_free_1 WHERE fclass='motorway'"
) %>% 
  st_transform(3083) %>%   # preemptively changing CRS to match our other data
  st_union()
```
Creating 200 meter buffer around Houston roads.
```{r}
# st_buffer with specified units finds all areas within that distance for us
highways_200m <-  st_buffer(highways, dist = units::set_units(200, "m")) %>% 
  st_union()
```

Combining highway data with its buffer
```{r}
highways <- st_simplify(highways, dTolerance = 10)
highways_200m <- st_simplify(highways_200m, dTolerance = 10)

highways_union <- st_union(highways, highways_200m)
# Now we have a single object that includes the highways and 200m around them
```

Now, we find areas that experiences a blackout, but 200m away from highways.
```{r}
blackout_no_highway <- st_difference(diff_final, highways_200m)
```

### Identify the number of homes likely impacted by blackouts

Importing houses data
```{r}
#| output: false
# Loading in houses vector data (multipolygon object)
houses <- st_read(here("data/gis_osm_buildings_a_free_1.gpkg"),
                 query = "SELECT *
FROM gis_osm_buildings_a_free_1
WHERE (type IS NULL AND name IS NULL)
OR type in ('residential', 'apartments', 'house', 'static_caravan', 'detached')") %>% 
  st_transform(3083)  # preemptively changing CRS to match our other data
```

We have to create a layer of homes that are only within our `blackout_no_highway` polygon. This is because we want to see *only* difference in light intensity greater than 200 and unaffected by highway light changes.
```{r}
# We use st_contains to find the houses that are fully contained by our blackout polygon
houses_blackout_sgbp <- st_contains(blackout_no_highway, houses)

houses_blackout_logical <- lengths(houses_blackout_sgbp) > 0 

houses_blackout <- houses[houses_blackout_logical, ]

# Ensuring that our cropping worked by checking for difference in cell number
if(ncell(blackout_no_highway) == ncell(houses_blackout)){
  warning("Failed to extract house data 200m away from highways")
} else {
  print("Successfully extracted only houses affected by blackouts (200m away from highways)")
}
```


### Blackout by Census Tracts

Add description here
```{r}
# Checking layers within our socioeconomic data
layers <-  st_layers(here("data/ACS_2019_5YR_TRACT_48_TEXAS.gdb"))

# Importing income and geometry layers
income <- st_read(here("data/ACS_2019_5YR_TRACT_48_TEXAS.gdb"), layer = "X19_INCOME")
census_tracts <- st_read(here("data/ACS_2019_5YR_TRACT_48_TEXAS.gdb"), layer = "ACS_2019_5YR_TRACT_48_TEXAS")
```
Joining the two data frames into one
```{r}
# census_tracts stores geo_ID under a different column names -- fixing it to match income
census_tracts <- census_tracts %>% 
  select(-GEOID) %>% 
  rename(GEOID = GEOID_Data)

# Combining the two layers, and adding geometry column
income_tracts <- left_join(census_tracts, income, by = "GEOID") %>% 
  mutate(shape = census_tracts$Shape) 
```

Attempt to create check -- revisit
```{r}
#| execute: false
# if(income_tracts$shape == census_tracts$Shape & 
#   income_tracts$GEOID == census_tracts$GEOID){
#  warning("Adding geometry column to combined data frame did not work")
#} else {
#  print("Successfully added shape column to data frame, and made it into an sf object")
#}
```

Combining house data with income/census tract data
```{r}
# Ensuring income_tracts has the same CRS
income_tracts <- st_transform(income_tracts, crs = st_crs(houses_blackout))

# Joining tract data with houses data
houses_tracts <- st_join(income_tracts, houses_blackout, join = st_within)

houses_tracts <- houses_tracts %>%
  st_drop_geometry() %>% 
  mutate(median_income = B19013e1)
```

```{r}
# transforming CRS' to match
census_tracts <- st_transform(census_tracts, crs = st_crs(houses_blackout))

# filtering census_tracts to only include tracts that contain houses w blackouts
census_blackout <- st_filter(census_tracts, houses_blackout)

# trying to find the opposite of census blackout DELETE
census_noblackout_sgbp <- st_disjoint(census_blackout, census_tracts)
census_noblackout_logical <- lengths(census_noblackout_sgbp) < 0
census_noblackout <- census_tracts[census_noblackout_logical,]

# Joining tract data to match census data that experienced blackouts DELETE?
houses_tracts <- st_join(income_tracts, census_blackout, join = st_within)

if(ncell(census_tracts) == ncell(census_blackout)){
  warning("Filtering did not remove cells")
} else {
  print("Filtering successfully removed cells")
}
```


## Maps, Figures and Reflection

The following section contains the following outputs:
- a set of maps comparing night light intensities before and after the first two storms
- a map of the homes in Houston that lost power
- an estimate of the number of homes in Houston that lost power
- a map of the census tracts in Houston that lost power
- a plot comparing the distributions of median household income for census tracts that did and did not experience blackouts
- a brief reflection (approx. 100 words) summarizing your results and discussing any limitations to this study

### Comparing Light Intensities (Map)

```{r}
# Cropping both days of our light data to just the Houston area
# Using previously made crop box
feb7_cropped <- st_crop(feb7_lights, crop_box)
feb16_cropped <- st_crop(feb16_lights, crop_box)

# Plotting light intensity for both days
feb7 <- plot(feb7_cropped,
     main = "Before Storm (2021-02-07)")

feb16 <- plot(feb16_cropped,
     main = "After Storm (2021-02-07)")
```

### Homes in Houston that Lost Power (Map & Esimate)

**Map**

```{r}
tm_shape(houses_tracts, bbox = houses_blackout) +
  tm_polygons(fill = "#c6b8cf",
              alpha = 0.25,
              border.col = "gray90",
              lwd = 0.75) +
tm_title(text = "Houston, TX: Homes that experienced blackout in February 2021") +
tm_compass() +
tm_scale_bar() +
tm_shape(houses_blackout) +
  tm_symbols(fill = "darkblue",
             border.col = "darkblue",
             alpha = 0.5,
             size = 0.15) 
```

**Estimate**

Our data frame `houses_blackout` only contains houses in areas that we determined experience a blackout. Therefore, to find the number of homes in Houston that experiences a blackout, we can simply count the number of rows in our data frame.
```{r}
houses_estimate <- nrow(houses_blackout)

print(paste("The number of homes that lost power in Houston between February 7th and 16th, 2021, is", houses_estimate, "."))
```

## Blackout and Census Tract
```{r}
tm_shape(census_tracts, bbox = houses_blackout) +
  tm_polygons(fill = "#c6b8cf",
              border.col = "gray90",
              lwd = 0.75) +
tm_shape(census_blackout) +
  tm_polygons(fill = "#482161",
              border.col = "#c6b8cf",
              lwd = 0.75) +
tm_add_legend(labels = c("Experienced blackout", "Did not experience blackout"),
              fill = c("#482161", "#c6b8cf"),
              title = "Census Blocks",
              is.portait = TRUE) +
tm_title("February 2021 Storms: Houston, TX") +
tm_compass() +
tm_scale_bar()
```

a plot comparing the distributions of median household income for census tracts that did and did not experience blackouts


```{r}
# census blocks that experiences blackout
# census_blackout

# all census blocks with income data
# income_tracts

# I don't need geometries anymore

income_tracts_df <- income_tracts %>% 
  st_drop_geometry() %>% 
  mutate(median_income = B19013e1)

census_blackout_df <- census_blackout %>% 
  st_drop_geometry()

# joining onto census blackout df will only leave income data for census blocks with blackouts
blackout_income <- inner_join(income_tracts_df, census_blackout_df, by = "NAMELSAD")

# to keep the opposite observations, we can use anti_join
noblackout_income <- anti_join(income_tracts_df, blackout_income, by = "NAMELSAD")

blackout_income %>% 
  summary()
```

```{r}
# plot of median income distribution of houses experiencing blackout
plot1 <- blackout_median <- ggplot(blackout_income, aes(y = median_income)) +
  geom_boxplot() +
labs(title = "Census Tracts Experiencing Blackout")

plot2 <- noblackout_median <- ggplot(noblackout_income, aes(y = median_income)) +
  geom_boxplot() +
labs(title = "Census Tracts No Blackout")

library(patchwork)

plot1 + plot2
```

Questions for Office Hours:
- how do I change color bar/legend as well as color in general for plots showing differences in light intensity
- ^ also, adding long/lat lines possible?
- is estimate just number of rows? and is plot of just dots ok? 
- are my distributions of median income right???