---
title: "Texas Power Crisis (February 2021)"
author: "Sofia Sarak"
format: html
editor: visual
---

## Initial Set Up

```{r}
#| messages: false
# Loading necessary libraries
library(tidyverse)
library(here)
library(terra)
library(sf)
library(tmap)
```

### Loading in and Cleaning Data

Night Lights
```{r}
#| output: false
# Loading in raster (night light) data
feb7_tile5 <- rast(here("data/VNP46A1/VNP46A1.A2021038.h08v05.001.2021039064328.tif"))
feb7_tile6 <- rast(here("data/VNP46A1/VNP46A1.A2021038.h08v06.001.2021039064329.tif"))
feb16_tile5 <- rast(here("data/VNP46A1/VNP46A1.A2021047.h08v05.001.2021048091106.tif"))
feb16_tile6 <- rast(here("data/VNP46A1/VNP46A1.A2021047.h08v06.001.2021048091105.tif"))
```

We will have to merge these 4 rasters into 2 (by day). 

```{r}
#| execute: false
# Merging rasters together, by day
feb7_lights <- c(feb7_tile5_ext, feb7_tile6_ext)
feb16_lights <- c(feb16_tile5_ext, feb16_tile6_ext)
```

Our first attempt tells us that the extent of tiles 5 and 6 does not match. We can fix that using the terra::extend function.
```{r}
# Checking if extents of rasters match
ext(feb7_tile5) == ext(feb7_tile6)
ext(feb16_tile5) == ext(feb16_tile6)

# Making tile 5 match the extent of tile 6
# Tile 6 contains all of Houston
feb7_tile5_ext <- extend(feb7_tile5, feb7_tile6)
feb16_tile5_ext <- extend(feb16_tile5, feb7_tile6)

# and vive versa
feb7_tile6_ext <- extend(feb7_tile6, feb7_tile5_ext)
feb16_tile6_ext <- extend(feb16_tile6, feb7_tile5_ext)

# Checking again if extents of rasters match
ext(feb7_tile5_ext) == ext(feb7_tile6_ext)
ext(feb16_tile5_ext) == ext(feb16_tile6_ext)
```
```{r}
# Merging rasters together, by day
feb7_lights <- c(feb7_tile5_ext, feb7_tile6_ext)
feb16_lights <- c(feb16_tile5_ext, feb16_tile6_ext)
```

Our raster tiles are merged, but now contain two layers -- one for each tile, as there was significant spatial overlap. To rectify this, we will combine our layers by taking the mean of each cell.

```{r}
# Assigning Feb 7th data frame to the mean of both layers (tiles 5 and 6)
feb7_lights <- mean(feb7_lights)
names(feb7_lights) <- "Light Intensity (nW cm^-2sr^-1" # changing name from "mean" to be more accurate

# Assigning Feb 16th data frame to the mean of both layers (tiles 5 and 6)
feb16_lights <- mean(feb16_lights)
names(feb16_lights) <- "Light Intensity (nW cm^-2sr^-1" # changing name from "mean" to be more accurate
```


Roads
```{r}
#| output: false
# Loading in roads vector data (line object)
roads <- st_read("data/gis_osm_roads_free_1.gpkg")
```
Houses
```{r}
#| output: false
# Loading in houses vector data (multipolygon object)
houses <- st_read(here("data/gis_osm_buildings_a_free_1.gpkg"))
```
Socioeconomic
```{r}
#| output: false
socioeconomic <- st_layers(here("data/ACS_2019_5YR_TRACT_48_TEXAS.gdb"))
```

### Data Exploration

Are the CRS' the same?
CHECK FOR SCIOECOMIC AS WELL
```{r}
st_crs(houses) == st_crs(roads)
st_crs(feb7_lights) == st_crs(feb16_lights)

st_crs(houses) == st_crs(feb7_lights)
```


## Creating Blackout Mask

We would like to explore only the areas of Houston that experienced a blackout. To do this, we can create a "mask" based on the difference in light intensity from February 7, 2021 and February 16, 2021.

```{r}
# Finding difference in light intensity
# Using raster algebra
diff_light <- feb7_lights - feb16_lights

# Reclassifying raster
# Creating reclassification matrix
rcl <- matrix(c(0, 200, 1, # group 1: drop in less than 200 nW cm^-2sr^-1 
                200,10000, 2, # group 2: drop in >= 200 nW cm^-2sr^-1 
                ncol = 3, byrow = TRUE
              ))
# Using reclass matrix to reclassfiy difference raster
diff_reclass <- classify(diff_light, rcl = rcl)

# Assign labels to the numerical categories
levels(diff_reclass) <- tibble::tibble(id = 1:2, 
                                  cats = c("No Blackout", "Blackout"))

summary(diff_reclass)
```




